---
title: "Data Manipulation"
author: "Sophy Jones"
date: "07/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the packages needed
```{r}
library(raster) # for dealing with spatial data
library(dplyr) # for handy data manipulation functions
library(tidyr) # ditto
library(lme4) # for mixed effects models
library(car) # for getting anova tables with significance values
library(DHARMa) # for model criticism plots
library(MuMIn) # for checking explanatory power of mixed effects models
library(sjPlot) # for visualising results
library(effects) # for extracting model effects
library(ggplot2) # for plotting
library(merTools) # useful for a few things, but we're using it for extracting estimates for plotting
```

Read in the data 

```{r}
diversity <- readRDS("UP-TO-DATE.rds") 
```

Create new data frame with just insect data

```{r}
insects <- droplevels(subset(diversity, Class=="Insecta")) 
```

### Correct for sampling effort across sites within a study

```{r}
studies_check <- diversity %>%
  # filter the rows where Sampling efforts are NA
  filter(is.na(Sampling_effort)) %>%
  # keep only unique studies
  distinct(SS) %>%
  # pull the vector
  pull(SS)

diversity %>%
  # filter the rows where studies are those that had missing sampling efforts (above)
  filter(SS %in% studies_check) %>%
  # drop missing levels
  droplevels() %>%
  # pull out the sampling efforts of these studies
  pull(Sampling_effort)%>%
  # summarise to check that ALL the data are NAs
  summary()

diversity <- diversity %>%
  
  # replace missing sampling effort values with 1
  mutate(x = replace_na(Sampling_effort, 1)) %>%
  
  # group by Study
  group_by(SS) %>%
  
  # check how many sampling efforts there are in each study
  mutate(n_sample_effort = n_distinct(Sampling_effort),
         # get the maximum sampling effort for the studies
         
         max_sample_effort = max(Sampling_effort)
         ) %>%
  
  ungroup() %>%
  
  # if the study has more than one sampling effort, correct the abundance

  # so if there's only one sampling effort, then create a 'dummy sampling effort' of 1 so that we don't change the abundances when we do the divisions. Otherwise, we give it the maximum sampling effort.
  mutate(DividingEffort = ifelse(n_sample_effort == 1, 1, max_sample_effort)) %>%
  
  # if the diversity metric isn't sensitive to the effort, then we'll change the value to 1 too (so we won't end up changing the measurement), otherwise leave it as it is
  mutate(DividingEffort = ifelse(Diversity_metric_is_effort_sensitive == FALSE, 1, DividingEffort)) %>%
  
  # now let's create the effort corrected measurement by dividing the abundances by the sampling efforts
  mutate(Corrected_sampling_effort = Sampling_effort / max_sample_effort,
         Effort_corrected_measurement = Measurement * Corrected_sampling_effort) 

# summarise the corrected sampling efforts
summary(diversity$Corrected_sampling_effort)
```

Test what we have done works
```{r}
diversity %>% 
  # filter out the test study
  filter(SS == "AD1_2001__Liow 1") %>% 
  # select out the columns to check
  dplyr::select(Measurement, Effort_corrected_measurement, Sampling_effort, Corrected_sampling_effort)

darvill_test <- diversity %>% 
  filter(SS == "AD1_2004__Darvill 1") %>% 
  dplyr::select(Measurement, Effort_corrected_measurement)

darvill_test
```

###  Merge any sites that are within the same land-use type and that have identical coordinates, start and end dates

```{r}
diversity <- diversity %>%
  
  # group by aspects of the sites that should be identical if we need to merge the abundances
  # I only want to merge abundances if they are within the same study and block
  # as I'm assuming that even if the locations and sampling times are the same, if the blocks or studies are different, then there is some good reason for this.
  group_by(Source_ID, Study_number, Study_name, Block,
           #diversity metric type
           Diversity_metric, Diversity_metric_type, Diversity_metric_unit,
           Diversity_metric_is_effort_sensitive,
           
           #details of the sites
           Predominant_habitat, Use_intensity, Years_since_fragmentation_or_conversion,
           
           #details of the sampling method
           Sampling_method, Sampling_effort_unit,
           
           #species identity
           Study_common_taxon, Rank_of_study_common_taxon,
           Taxon_number, Taxon_name_entered,
           Indication, Parsed_name,
           Best_guess_binomial, COL_ID, Taxon, Name_status,
           Rank, Kingdom, Phylum, Class, Order, Family, Genus, Species,
           Higher_taxon,
           
           #site location
           Longitude, Latitude,
           
           #sampling time
           Sample_start_earliest, Sample_end_latest, Sample_date_resolution) %>%
  
  # if the diversity metric is occurrence:
  #   if it is present at all, give it a 1, if it is always absent, give it a 0,
  # otherwise (if the metric is either abundance or species richness):
  #   calculate the weighted abundance/richness for each taxonomic group, weighted by sampling effort
  
  mutate(merged_diversity = 
           ifelse(Diversity_metric_type == "Occurrence",
                  # if any of the occurrence values are 1, `any` will return TRUE. If you sum a logical, TRUE becomes 1 and FALSE becomes 0
                  sum(any(Effort_corrected_measurement > 0)),
                  
                  # note that since we've already corrected the sampling effort, this is essentially a mean rather than a weighted mean for abundance measurements. It's a weighted mean for species richness though where sampling efforts vary.
                  stats::weighted.mean(x = Effort_corrected_measurement,
                                       w = Corrected_sampling_effort))
         )

# pull out the grouping data (so we can double check how many records we're merging for each)
group_dat <- diversity %>% 
  group_data() %>%
  mutate(nvals_merged = lengths(.rows),
         merge_ID = row_number())

# ungroup the diversity data for future use
diversity <- ungroup(diversity)

# create a dataset where we can extract just the merged data if we want to
diversity_merged <- diversity %>%
  left_join(group_dat)
```

Check that what we have done has worked - row numbers should be equal right now

```{r}
nrow(diversity) == nrow(diversity_merged)
```

More tests to make sure it has worked

```{r}
test_data <- diversity_merged %>%
  filter(nvals_merged > 1) %>%
  distinct(merge_ID, .keep_all = TRUE)

test1 <- diversity_merged %>%
  filter(merge_ID == test_data$merge_ID[1]) %>%
  dplyr::select(SS, SSB,
         Diversity_metric, Diversity_metric_type, Diversity_metric_unit,
         Predominant_habitat, Use_intensity, Years_since_fragmentation_or_conversion,
         Sampling_method, Sampling_effort_unit,
         Study_common_taxon, Rank_of_study_common_taxon,
         Taxon_name_entered,
         Best_guess_binomial,
         Longitude, Latitude,
         Sample_start_earliest, Sample_end_latest, Sample_date_resolution,
         Effort_corrected_measurement,
         Corrected_sampling_effort,
         merged_diversity,
         .rows,
         nvals_merged,
         merge_ID
         )

test1

paste(test_data$SS[8000])
```

```{r}
test2 <- diversity_merged %>%
  filter(merge_ID == test_data$merge_ID[8000]) %>%
  dplyr::select(SS, SSB,
         Diversity_metric, Diversity_metric_type, Diversity_metric_unit,
         Predominant_habitat, Use_intensity, Years_since_fragmentation_or_conversion,
         Sampling_method, Sampling_effort_unit,
         Study_common_taxon, Rank_of_study_common_taxon,
         Taxon_name_entered,
         Best_guess_binomial,
         Longitude, Latitude,
         Sample_start_earliest, Sample_end_latest, Sample_date_resolution,
         Effort_corrected_measurement,
         Corrected_sampling_effort,
         merged_diversity,
         .rows,
         nvals_merged,
         merge_ID
         )

test2
```

Check studies that didn't need merging

```{r}
test_data <- diversity_merged %>%
  filter(nvals_merged == 1) %>%
  distinct(merge_ID, .keep_all = TRUE)

paste(test_data$SS[1])


test3 <- diversity_merged %>%
  filter(merge_ID == test_data$merge_ID[1]) %>%
  dplyr::select(SS, SSB,
         Diversity_metric, Diversity_metric_type, Diversity_metric_unit,
         Predominant_habitat, Use_intensity, Years_since_fragmentation_or_conversion,
         Sampling_method, Sampling_effort_unit,
         Study_common_taxon, Rank_of_study_common_taxon,
         Taxon_name_entered,
         Best_guess_binomial,
         Longitude, Latitude,
         Sample_start_earliest, Sample_end_latest, Sample_date_resolution,
         Effort_corrected_measurement,
         Corrected_sampling_effort,
         merged_diversity,
         .rows,
         nvals_merged,
         merge_ID
         ) 

test3
```


```{r}

```

